Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula

### 2024-03-10 output/paper/manuscript.qmd
Started reviewing manuscript based on co-authors comments.

---
title: "Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula"

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention measures in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of efficiency from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattentiveness outcomes in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

One group of {{< acr edcs >}} that may have a deleterious effect on neurodevelopment is the {{< acr opps >}}, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has been associated with several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to {{< acr bpa >}} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

@@

Taken together, these results suggest that the negative influence of exposure to certain {{< acr edcs >}} on neurodevelopmental outcomes might be mediated, at least partially, by disruption of the {{< acr hpa >}} axis' homeostasis.
In the present study, we thus estimated associations between 1) non-persistent {{< acr edcs >}} and attention, 2) non-persistent {{< acr edcs >}} and glucocorticosteroids, and 3) glucocorticosteroids and attention, using the parametric g-formula and {{< acr mcs first_use=true >}}, in children of a large cohort in Europe.

@@

### Neurodevelopment {#sec-neurodevelopment}

Neurodevelopmental outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the {{< acr ant first_use=true >}} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency in three different components of attention: alerting, orienting, and executive attention. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found in @MaitreBontCasas:2018. The outcome of interest for the present study is the {{< acr hitrtse first_use=true >}} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high {{< acr hitrtse >}} indicates highly variable reaction times, and is considered a measure of inattentiveness.

@@

To reduce the effect of measured confounders on the exposure-outcome association, stabilized balancing weights were estimated using the energy method available in the `WeightIt` R package [@Greifer:2023]. This methods estimates weights by minimizing an energy statistic related to covariate balance [@HulingGreiferChen:2023], thus avoiding the need to specify a parametric model.
Weights below the 0.1 and above the 0.9 quantiles were trimmed. Trimming might lead to decreased covariate balance and potentially change the estimand, but can also decrease the variability of the weights.
Covariate balance was assessed using functionalities provided by the `cobalt` R package [@Greifer:2023a]. Specifically, we used *Love* plots to visualize covariate balance before and after adjusting.

@@

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} was associated with {{< acr hitrtse >}}, a measure of inattentiveness, and with total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr cis >}} included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to {{< acr hitrtse >}}. The literature on non-persistent {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of non-persistent {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].

Our results are consistent with prior epidemiological research that associated exposure to certain non-persistent {{< acr edcs >}} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixture. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and {{< acr opps >}}, in relation to urinary glucocorticosteroids in childhood. Nonetheless, previous toxicological studies provide evidence for the inhibition by phthalates of human 11$\beta$-hydroxysteroid dehydrogenase 2 (11$\beta$-HSD2) activity, responsible for the conversion of active cortisol into inactive cortisone [@ZhaoChuHuang:2010; @MaLianDong:2011]. There is also *in silico* evidence suggesting that {{< acr bpa >}}, a phenol, and Triazophos (TAP), a organophosphorus insecticide, can bind to the human glucocorticoid receptor [@PrasanthDivyaSadasivan:2010; @YangLiRen:2019].

We are also not aware of prior epidemiological studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to {{< acr hitrtse >}}, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain [@LupienMcEwenGunnar:2009]. While we did find sex-specific evidence of an effect, their clinical relevance is questionable.

@@

In conclusion, in a study of 1,297 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent {{< acr edcs >}} in childhood might have short-term effects on {{< acr hitrtse >}} in childhood, (ii) exposure to non-persistent {{< acr edcs >}} in childhood might disrupt the {{< acr hpa >}} axis in childhood, and (iii) disruption of the {{< acr hpa >}} axis in childhood might have short-term, sex-specific effects on {{< acr hitrtse >}}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to non-persistent {{< acr edcs >}} on childhood neurodevelopment in larger populations.

>>>>

---
title: "Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and attentional function: A study based on the parametric g-formula"

@@

Evidence suggests that {{< acr edcs >}} may perturb the {{< acr hpa >}} axis, which has a major role in brain development. We aimed to evaluate the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and inattention in children using data from the {{< acr helix >}} cohort. We used the parametric g-formula to estimate effects between {{< acr edcs >}}, glucocorticosteroids, and {{< acr hitrtse >}}, a measure of inattention from the {{< acr ant >}}, and tested for possible effect modification by sex. We observed a positive {{< acr mc >}} for exposure increases from the 10th to the 90th percentile for methyl-paraben (`r format_res(meffects1, "mepa", "HRT-SE")`), and the phthalate metabolites oxo-MiNP (`r format_res(meffects1, "oxominp", "HRT-SE")`), oh-MiNP (`r format_res(meffects1, "ohminp", "HRT-SE")`), and MEHP (`r format_res(meffects1, "mehp", "HRT-SE")`), on {{< acr hitrtse >}}. Several {{< acr edcs >}} were also associated with a positive {{< acr mc >}} for cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on {{< acr hitrtse >}}, although we found a possible effect modification by sex. Our results suggest that multiple {{< acr edcs >}} might interfere with inattention in children and with the homeostasis of the {{< acr hpa >}} axis.

@@

One group of {{< acr edcs >}} that may have a deleterious effect on neurodevelopment is the {{< acr opps >}}, although the few studies assessing exposure during childhood and through the use of biomarkers suffered from a series of limitations, including a small sample size [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. Exposure to phthalates and their metabolites during childhood and early adolescence has also been associated with several adverse neurodevelopmental outcomes, but these studies were limited to few phthalate metabolites and small study populations [@RamirezGalvez-OntiverosGonzalez-Domenech:2022]. The effects of exposure to {{< acr bpa >}} during childhood on cognitive functions are still unclear [@RamirezGalvez-OntiverosGonzalez-Domenech:2022].

@@

Taken together, these results suggest that the negative influence of exposure to certain {{< acr edcs >}} on neurodevelopmental outcomes might be mediated, at least partially, by disruption of the {{< acr hpa >}} axis' homeostasis.
In the present study, we thus estimated cross-sectional associations between 1) non-persistent {{< acr edcs >}} and attentional function, 2) non-persistent {{< acr edcs >}} and glucocorticosteroids, and 3) glucocorticosteroids and attentional function, using the parametric g-formula and {{< acr mcs first_use=true >}}, in children of a large network of cohorts in Europe.

@@

### Attentionall function {#sec-neurodevelopment}

Cognitive and motor function outcomes were assessed with standardized, non-linguistic, and culturally blind computer tests, including the {{< acr ant first_use=true >}} [@RuedaFanMcCandliss:2004], which provides a measure of efficiency of attentional function. The tests were administered in a standardized way, and with minimal interference from the field workers. Further information can be found in @MaitreBontCasas:2018. The outcome of interest for the present study is the {{< acr hitrtse first_use=true >}} [@SunyerEsnaolaAlvarez-Pedrerol:2015], a measure of response speed consistency throughout the test. A high {{< acr hitrtse >}} indicates highly variable reaction times, and is considered a measure of inattentiveness.

@@

To reduce the effect of measured confounders on the exposure-outcome association, stabilized balancing weights were estimated using the energy method available in the `WeightIt` R package [@Greifer:2023]. This method estimates weights by minimizing an energy statistic related to covariate balance [@HulingGreiferChen:2023], thus avoiding the need to specify a parametric model.
Weights below the 0.1 and above the 0.9 quantiles were trimmed. Trimming might lead to decreased covariate balance and potentially change the estimand, but can also decrease the variability of the weights.
Covariate balance was assessed using functionalities provided by the `cobalt` R package [@Greifer:2023a]. Specifically, we used *Love* plots to visualize covariate balance before and after adjusting.

@@

The impact of exposure to {{< acr edcs >}} on human health has attracted considerable research interest. While research in this area has mainly investigated the effects of prenatal exposure on child neurodevelopment [@RamirezGalvez-OntiverosGonzalez-Domenech:2022], little is still known about childhood exposure. In this study, consisting of 1,297 children from 6 European birth cohorts, we observed that short-term childhood exposure to non-persistent {{< acr edcs >}} was associated with attentional function, and with total production of cortisol, cortisone, and corticosterone, although the majority of the {{< acr cis >}} included the null effect. Increased production of these glucocorticosteroids did not seem to affect {{< acr hitrtse >}}. Some of these effects differed for females and males, including significant differences for the effects of increased production of cortisone and corticosterone on {{< acr hitrtse >}}. Specifically, an increased production of these glucocorticosteroids was associated with lower values of {{< acr hitrtse >}} for females, and higher values for males. Taken together, these results suggest that these non-persistent {{< acr edcs >}} might be responsible for perturbations of the {{< acr hpa >}} axis' homeostasis, and that higher levels of these glucocorticosteroids might interfere with different functions of attention in a sex-specific manner.

We are not aware of prior studies specifically investigating the effects of exposure to {{< acr edcs >}} in relation to attentional function. The literature on non-persistent {{< acr edcs >}} and neurodevelopment in children has mostly focused on {{< acr opps >}} [@BouchardBellingerWright:2010; @Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @CartierWarembourgLeManer-Idrissi:2016; @YuDuChiou:2016], phthalate metabolites [@HuangChenSu:2015; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023], and {{< acr bpa >}} [@TewarAuingerBraun:2016; @LiZhangKuang:2018; @Rodriguez-CarrilloMustielesPerez-Lobato:2019], in relation to {{< acr adhd >}} [@BouchardBellingerWright:2010; @TewarAuingerBraun:2016; @YuDuChiou:2016; @LiZhangKuang:2018], and intelligence scales [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @HuangChenSu:2015; @CartierWarembourgLeManer-Idrissi:2016; @HuangTsaiChen:2017; @KimHongShin:2017; @VilmandBeckBilenberg:2023]. Few studies have looked into different classes of non-persistent {{< acr edcs >}} [@ShoaffCoullWeuve:2020 in relation with the Conners Attention Deficit Scale and the Behavior Assessment System for Children; @OhKimKannan:2023 in relation with {{< acr adhd >}} symptoms]. Overall, and consistent with our results, these studies seem to provide further evidence of the adverse effects of several non-persistent {{< acr edcs >}} on neurodevelopment (too broad) in children. While not all these studies have investigated effect modification by sex, it seems that these adverse effects tend to be stronger in males [@Gonzalez-AlzagaHernandezRodriguez-Barranco:2015; @YuDuChiou:2016; @ShoaffCoullWeuve:2020], although this difference might depend on the timing of exposure [@].

Our results are consistent with prior epidemiological research that associated exposure to certain non-persistent {{< acr edcs >}} with higher levels of cortisol [@KimLeeMoon:2018; @SunLiJin:2018; @SearsLiuLanphear:2023]. There are some differences, though. First, these studies only focus on phthalates, either as individual metabolites or as mixtures. Second, exposure assessment in @SunLiJin:2018 and @KimLeeMoon:2018 was performed during gestation or the first 15 months of life, respectively. Finally, the glucocorticosteroids were measured in cord blood [@SunLiJin:2018] and hair [@SearsLiuLanphear:2023]. Contrary to these studies, we did find effect modification by sex. We are not aware of other epidemiological studies investigating phthalates metabolites, phenols, and {{< acr opps >}}, in relation to urinary glucocorticosteroids in childhood. Nonetheless, previous toxicological studies provide evidence for the inhibition by phthalates of human 11$\beta$-hydroxysteroid dehydrogenase 2 (11$\beta$-HSD2) activity, responsible for the conversion of active cortisol into inactive cortisone [@ZhaoChuHuang:2010; @MaLianDong:2011]. There is also *in silico* evidence suggesting that {{< acr bpa >}}, a phenol, and Triazophos (TAP), a organophosphorus insecticide, can bind to the human glucocorticoid receptor [@PrasanthDivyaSadasivan:2010; @YangLiRen:2019].

We are also not aware of prior epidemiological studies specifically investigating the effects of elevated levels of glucocorticosteroids in relation to attentional function, although there is evidence that under- or over-production of glucocorticosteroids interfere with the normal development of the brain [@LupienMcEwenGunnar:2009]. While we did find sex-specific evidence of an effect, their clinical relevance is questionable.

@@

In conclusion, in a study of 1,297 children from 6 European birth cohorts, we observed that (i) exposure to non-persistent {{< acr edcs >}} in childhood might have short-term effects on {{< acr hitrtse >}} in childhood, (ii) exposure to non-persistent {{< acr edcs >}} in childhood might disrupt the {{< acr hpa >}} axis in childhood, and (iii) disruption of the {{< acr hpa >}} axis in childhood might have short-term, sex-specific effects on {{< acr hitrtse >}}. Future studies should investigate how glucocorticosteroids might mediate the adverse effects of exposure to non-persistent {{< acr edcs >}} on childhood neurodevelopment (too broad) in larger populations.
